---
title: "Practical Machine Learning Course Project"
author: "Gerardo Mondragón"
date: "8/11/2020"
output: 
    html_document:
        keep_md: true
---

# **Overview**
This is the write-up part of the course project. The data is about a group of people doing barbell lifts correctly and incorrectly, as measured by on-body accelerometers. The goal is to predict, from this data, if the lift was made correctly or incorrectly. The source for this data is:

Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. **Qualitative Activity Recognition of Weight Lifting Exercises**. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013.

# **Setting up the environment**
## **Loading libraries**

```{r setup, warning=FALSE, message=FALSE}
library(caret)
library(rpart)
library(rpart.plot)
library(rattle)
library(randomForest)
library(corrplot)
set.seed(123)
```

## **Loading and cleaning up the data**
```{r load data, cache=TRUE}
# Download and read the data
training <- read.csv(url("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"))
testing <- read.csv(url("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"))

# Partition the data
inTrain <- createDataPartition(training$classe, p = 0.7, list = F)
trainset <- training[inTrain, ]
testset <- training[-inTrain, ]
```

In order to reduce the amount of variables (160) in the partitions, we will remove such variables with near zero variance.
```{r remove NZV,cache=TRUE}
NZV <- nearZeroVar(trainset)
trainset <- trainset[, -NZV]
testset <- testset[, -NZV]
```

Now we only have 104 variables. However, we can further reduce this number by removing variables that are over 95% missing values. 
```{r remove NA}
isNA <- sapply(trainset, function(x) mean(is.na(x))) > 0.95
trainset <- trainset[, isNA == F]
testset <- testset[, isNA == F]
```

This results in 59 variables. Lastly, we will remove the first five columns as they are identification variables, and so at the end, we have 54 variables.
```{r remove first cols}
trainset <- trainset[, -(1:5)]
testset <- testset[, -(1:5)]
```

Now we need to set the *classe* variable to be a factor instead of a character variable.
```{r factor classe}
trainset$classe <- as.factor(trainset$classe)
testset$classe <- as.factor(testset$classe)
```

# **Analysis**

The first step to my analysis is a correlation analysis before any actual modeling. The most strongly correlated variables appear in darker colors in the plot shown below. 
```{r correlation}
corrMatrix <- cor(trainset[, -54])
corrplot(corrMatrix, order = "FPC", method = "color", type = "lower",
         tl.cex = 0.8, tl.col = rgb(0, 0, 0))
```

# **Prediction Model**
Now for the main part, fitting a model that best predicts the way the barlifts were made. To this end, I will use three different models, and the one with highest accuracy will be deemed the best one. These models are:

- Random Forests
- Decision Tree
- Generalized Boosted Model (GBM)

## **Random Forests**
```{r random forests, cache=TRUE}
set.seed(123)
# Using trainControl() to control computational nuances.
RFcontrol <- trainControl("cv", 3, verboseIter = F)
# Fitting the model
fit_rf <- train(classe ~., method = "rf", data = trainset, trcontrol = RFcontrol)
fit_rf$finalModel
```

Now we create the prediction, create a confusion matrix and plot it.
```{r prediction rf}
predict_rf <- predict(fit_rf, newdata = testset)
confusionMat_rf <- confusionMatrix(predict_rf, testset$classe)
confusionMat_rf
plot(confusionMat_rf$table, col = confusionMat_rf$byClass,
     main = paste("Random Forest Accuracy =", round(confusionMat_rf$overall["Accuracy"], 4)))
```

## **Decision Trees**
```{r decision trees, warning=FALSE}
set.seed(123)
fit_dt <- rpart(classe ~., method = "class", data = trainset)
fancyRpartPlot(fit_dt)
```

Now we create the prediction, create a confusion matrix and plot it.
```{r prediction dt}
predict_dt <- predict(fit_dt, newdata = testset, type = "class")
confusionMat_dt <- confusionMatrix(predict_dt, testset$classe)
confusionMat_dt
plot(confusionMat_dt$table, col = confusionMat_dt$byClass,
    main = paste("Decision Trees Accuracy =", round(confusionMat_dt$overall["Accuracy"], 4)))
```

## **Generalized Boosted Model (GBM)**
```{r gbm, cache=TRUE}
set.seed(123)
GBMcontrol <- trainControl("repeatedcv", 5, 1)
fit_gbm <- train(classe ~., method = "gbm", data = trainset, trControl = GBMcontrol, verbose = F)
fit_gbm$finalModel
```

Now we create the prediction, create a confusion matrix and plot it.
```{r prediction gbm}
predict_gbm <- predict(fit_gbm, newdata = testset)
confusionMat_gbm <- confusionMatrix(predict_gbm, testset$classe)
confusionMat_gbm
plot(confusionMat_gbm$table, col = confusionMat_gbm$byClass, 
    main = paste("GBM Accuracy =", round(confusionMat_gbm$overall["Accuracy"], 4)))
```

# Predicting 
Since the best model as measured by accuracy is Random Forests, we will use this model to predict the values.
```{r predict testing}
results <- predict(fit_rf, newdata = testing)
results
```
